---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ohsome

<!-- badges: start -->
<!-- badges: end -->

The goal of ohsome is to ...

## Installation

If you have access to the repository and a 
[personal access token](https://docs.gitlab.com/ee/user/profile/personal_access_tokens.html){target=blank},
you can install the development version of ohsome from gitlab:

```{r, eval = FALSE}
remotes::install_gitlab(
	repo = "giscience/big-data/ohsome/libs/ohsome-r@develop",
	host = "https://gitlab.gistools.geog.uni-heidelberg.de",
	auth_token = personal_access_token
)

```

## Getting started

Upon attaching the *ohsome* package, a metadata request is sent to the *ohsome* 
API. The package message provides some essential metadata information, such as
the current temporal extent of the underlying OSHDB:

```{r library}
library(ohsome)
```

The metadata is stored in *.ohsome_metadata*. You can print it to the console
to get more details:

```{r metadata}
.ohsome_metadata
```

### Aggregate OpenStreetMap elements

### Other queries

### Bounding geometries
