---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-", 
  out.width = 900
)
```

# ohsome: An R package to interact with the ohsome API for OpenStreetMap history data aggregation and extraction

<!-- badges: start -->
[![LICENSE](https://img.shields.io/github/license/GIScience/ohsome-r)](LICENSE.md)
[![status: experimental](https://github.com/GIScience/badges/raw/master/status/experimental.svg)](https://github.com/GIScience/badges#experimental)
<!-- badges: end -->

This ohsome R package grants access to the power of the [ohsome API](https://api.ohsome.org){target=blank}
from R. ohsome lets you analyze the rich data source of the
[OpenStreetMap](https://www.openstreetmap.org/){target=blank} 
(OSM) history. It aims to leverage the tools of the 
[OpenStreetMap History Database](https://github.com/GIScience/oshdb){target=blank}
(OSHDB).

With ohsome, you can ...

- Get **aggregated statistics** on the evolution of OpenStreetMap elements and
specify your own temporal, spatial and/or thematic filters. The data aggregation 
endpoint allows you to access functions, e.g., to calculate the area of 
buildings or the length of streets at any given timestamp.

- Retrieve the **geometry** of the historical OpenStreetMap data, e.g., to
visualize the evolution of certain OpenStreetMap elements over time. You can get
the geometries for specific points in time or all changes within a timespan 
(full-history).

 
## Installation

You can install ohsome from [GitHub](https://github.com/GIScience/ohsome-r){target=blank}:

```{r, eval = FALSE}
remotes::install_github("GIScience/ohsome-r", ref = "dev-0.2.0")

```

## Getting started

Upon attaching the ohsome package, a metadata request is sent to the ohsome 
API. The package message provides some essential metadata information, such as
the current temporal extent of the underlying OSHDB:

```{r library}
library(ohsome)
```

The metadata is stored in `ohsome_metadata`. You can print it to the console
to get more details.

You can create any ohsome API query using the generic `ohsome_query()` function. 
It takes the endpoint path and any query parameters as inputs. For information 
on all available endpoints with their parameters, consult the 
[ohsome API documentation](https://docs.ohsome.org/ohsome-api/stable/endpoints.html){target=blank} 
or print `ohsome_endpoints` to the console.

However, this ohsome R package provides specific wrapper functions for queries 
to all available endpoints.

### OSM elements

#### Aggregating OSM elements

The
[elements aggregation endpoints](https://docs.ohsome.org/ohsome-api/stable/endpoints.html#elements-aggregation){target=blank}
of the ohsome API allow querying for the aggregated amount, length, area or 
perimeter of OpenStreetMap elements with given properties, within given 
boundaries and at given points in time.

Let us create a query for the total amount of breweries on OSM in the region of 
Franconia. The first argument to `ohsome_elements_count()` is the `sf` object 
`franconia` that is included in the 
[mapview](https://r-spatial.github.io/mapview/){target=blank} 
package and contains boundary 
polygons of the `r nrow(mapview::franconia)` districts of the region:

```{r fix, include = FALSE}
# avoid messages when handling franconia with old-style crs object
franconia <- sf::st_set_crs(mapview::franconia, 4326)
```

```{r elements_count}
library(mapview)

q <- ohsome_elements_count(franconia, filter = "craft=brewery")
```

The resulting `ohsome_query` object can be sent to the ohsome API with
`ohsome_post()`. By default, `ohsome_post()` returns the parsed API
response. In this case, this is a simple `data.frame` of only one row.

```{r post}
ohsome_post(q)
``` 

The `ohsome_query` object was created without an explicit `time` parameter. When
using functions to create element aggregation queries, `time` defaults to the 
most recent available timestamp in the underlying OSHDB.

Defining the `time` parameter unlocks the full power of ohsome API by giving 
access to the OSM history. It requires one or more 
[ISO-8601 conform timestring(s)](https://docs.ohsome.org/ohsome-api/stable/time.html){target=blank}.
Here is how to create a query for the number of breweries at the first of each 
month between 2010 and 2020:

```{r time, eval = FALSE}
ohsome_elements_count(franconia, filter = "craft=brewery", time = "2010/2020/P1M")
```

Alternatively, we can update the existing `ohsome_query` object `q` with the 
`set_time()` function, 
pipe ^[Instead of the new R native pipe `|>` you may choose to use `magrittr`'s `%>%`.]
the modified query directly into `ohsome_post()` 
and make a quick visualization with `ggplot2`:

```{r pipe, dev = "svg"}
library(ggplot2)

q |> 
	set_time("2010/2020/P1M") |>
	ohsome_post() |>
	ggplot(aes(x = timestamp, y = value)) +
	geom_line()
```

This is how to query the total number of breweries in all of Franconia. But
what if we want to aggregate the amount per district? The ohsome API provides 
specific endpoints for different grouped calculations, such as aggregation 
grouped by bounding geometry.

There are several ways to define a query for an aggregation grouped by boundary:

The `set_endpoint`function is used to change or append the endpoint path of an 
API request. In this case, we could append `groupBy/boundary` to the existing 
query to the `elements/count` endpoint. The endpoint path can either be given 
as a single string (`/groupBy/boundary`) or as a character vector: 
`set_endpoint(q, c("groupBy", "boundary"), append = TRUE)` 
^[The order of the elements in the character vector is critical!].

More comfortably, however, is the use of either the grouping argument with an
elements aggregation function (e.g. 
`ohsome_elements_count(grouping = "boundary)`) or of the `set_grouping()` 
function to modify an existing query object:

```{r groupBy_boundary, message = FALSE}
library(dplyr)

franconia |> 
	mutate(id = NAME_ASCI) |>
	ohsome_elements_count(filter = "craft=brewery", time = "2021-06-01") |>
	set_grouping("boundary") |>
	ohsome_post()
```

If you want your own identifiers for the geometries returned by ohsome, your 
input `sf` object needs a column explicitly named `id`. You can use `mutate()` 
or `rename()` from the 
[dplyr](https://dplyr.tidyverse.org){target=blank} 
package to create such a column as in the example
below.

By default, `ohsome_post()` returns an `sf` object whenever the ohsome API 
is capable of delivering GeoJSON data. This is the case for elements 
extraction queries as well as for aggregations grouped by boundaries.

Thus, you can easily create a choropleth map from the query results.
In addition, you can set the argument `return_value` to `density`. This will 
modify the endpoint path of the query so that ohsome return the number of 
breweries per area instead of the absolute value:

```{r density}
franconia |> 
	mutate(id = NAME_ASCI) |>
	ohsome_elements_count(filter = "craft=brewery", return_value = "density") |>
	set_time("2021-06-01") |>
	set_grouping("boundary") |>
	ohsome_post() |>
	mapview(zcol = "value", layer.name = "Breweries per sqkm")
```

#### Extracting OSM elements

The
[elements extraction endpoints](https://docs.ohsome.org/ohsome-api/stable/endpoints.html#elements-extraction){target=blank}
of the ohsome API allow obtaining geometries, bounding boxes or centroids of OSM
elements with given properties, within given boundaries and at given points in 
time. Together with the elements, you can choose to query for their tags and/or 
their metadata such as the changeset ID, the time of the last edit or the 
version number.

The following query extracts the geometries of buildings within 1000 m of 
Heidelberg main station with their tags. The response is used to visualize
the buildings and the values of their `building:levels` tag (if available):

```{r building_levels, warning = FALSE}
ohsome_elements_geometry(
	boundary = "8.67542,49.40347,1000", 
	filter = "building=* and type:way", 
	time = "2021-12-01",
	properties = "tags", 
	clipGeometry = FALSE
) |>
	ohsome_post() |>
	transmute(level = factor(`building:levels`)) |>
	mapview(zcol = "level", lwd = 0, layer.name = "Building level")
```

Similarly, you can use `ohsome_elements_centroid()` to extract centroids of OSM
elements and `ohsome_elements_bbox()` for their bounding boxes. Note that OSM
node elements (with point geometries) are omitted from the results if querying 
for bounding boxes.


#### Extracting the full history of OSM elements

While the elements extraction endpoints provide geometries and properties of OSM 
elements at specific timestamps, results of queries to the 
[full history endpoints](https://docs.ohsome.org/ohsome-api/v1/endpoints.html#elements-full-history-extraction){target=blank}
will include all changes to matching OSM features with corresponding 
`validFrom` and `validTo` timestamps.

Here, we request the full history of OSM buildings for the district of Schweinfurt
City, filter for features that still exist and visualize all building features 
with their year of creation:

```{r buildings}
meta <- ohsome_get_metadata()
start <- as.Date(meta$extractRegion$temporalExtent[1])
end <- as.Date(meta$extractRegion$temporalExtent[2])

schweinfurt <- franconia |> filter(NAME_ASCI == "Schweinfurt, Kreisfreie Stadt")

m <- schweinfurt %>%
	ohsome_elementsFullHistory_geometry(
		time = c(start, end),
		filter = "building=* and geometry:polygon", 
		clipGeometry = FALSE,
		properties = "metadata"
	) |> 
	ohsome_post() |>
	janitor:: clean_names() |>
	group_by(osm_id) |>
	mutate(year = min(format(valid_from, "%Y"))) |>
	filter(valid_to == end) |>
	mapview(zcol = "year", lwd = 0, layer.name = "Year of Feature Creation")

m@map %>% leaflet::setView(10.23, 50.04, zoom = 13)
```

You may find using `clean_names()` from the 
[janitor](https://github.com/sfirke/janitor){target=blank} 
package helpful in order
to remove special characters from column names in the parsed ohsome API 
response -- just as in the example above.

### OSM contributions

#### Aggregating OSM contributions

With queries to the ohsome API's
[contributions aggregation endpoints](https://docs.ohsome.org/ohsome-api/v1/endpoints.html#contributions-aggregation){target=blank},
you can get counts of the contributions provided by users to OSM. The following 
code requests the number of deletions of man-made objects at the location 
of the hypothetical 
[Null Island](https://en.wikipedia.org/wiki/Null_Island){target=blank}
per year between 2010 and 2020: 

```{r contribution_count}
ohsome_contributions_count(
	boundary = "0,0,10", 
	filter = "man_made=*", 
	time = "2010/2020/P1Y",
	contributionType = "deletion"
) |>
	ohsome_post()
```

The `contributionType` parameter is used to filter for specific types of 
contributions (in this case: deletions). If it is not set, any contribution is 
counted. Note that the resulting values apply to time intervals defined by a 
`fromTimestamp` and a `toTimestamp`.

#### Extracting OSM contributions

The 
[contributions extraction](https://docs.ohsome.org/ohsome-api/v1/endpoints.html#contributions-extraction){target=blank}
endpoints of the ohsome API can be used to extract feature geometries of 
contributions.

In the following example, we extract the centroids all amenities in the Berlin 
city district of Neukölln that have had contributions in March 2020. 
Consequently, we filter for features that have had tags changed and visualize 
their locations:

```{r contribution_extraction}
nominatimlite::geo_lite_sf("Berlin Neukölln", points_only = FALSE) |>
	ohsome_contributions_centroid() |>
	set_filter("amenity=*") |>
	set_time("2020-03,2020-04") |>
	set_properties("contributionTypes") |> 
	ohsome_post() |>
	filter(`@tagChange`) |>
	mapview(layer.name = "Amenities with Tag Changes")
```

### OSM users

You can get statistics on the number of users editing specific features through
the 
[users aggregation](https://docs.ohsome.org/ohsome-api/v1/endpoints.html#users-aggregation){target=blank}
endpoints of the ohsome API.

Here, we show the number of users editing buildings before, during and after 
Nepal earthquake 2015:

```{r nepal}
ohsome_users_count(
	boundary = "82.3055,6.7576,87.4663,28.7025",
	filter = "building=* and geometry:polygon",
	time = "2015-03-01/2015-08-01/P1M"
) |>
	ohsome_post()
```

### Bounding geometries

The ohsome API requires bounding geometries either as bounding polygons 
(`bpolys`), bounding boxes (`bboxes`) or bounding circles (`bcircles`) 
parameters to the query in a textual form (see
[ohsome API documentation](https://docs.ohsome.org/ohsome-api/stable/boundaries.html){target=blank}).
The ohsome R package uses the generic function `ohsome_boundary()` under the 
hood to make your life easier. It accepts a wider range of input geometry 
formats, while guessing the right type of bounding geometry.

As seen above, `sf` objects can be passed into the `boundary` argument of 
`ohsome_query()` and any of its wrapper functions. You can also update queries 
with `set_boundary()`. The `sf` object will be converted into GeoJSON and passed 
into the `bpolys` parameter of the query.

If you wish to aggregate or extract OSM elements on administrative boundaries in
the `sf` format, you might want to check out packages such as 
[rnaturalearth](https://github.com/ropensci/rnaturalearth){target=blank}, 
[geodata](https://github.com/rspatial/geodata){target=blank},
[raster](https://github.com/rspatial/raster){target=blank} 
(in particular its `getData()` function), 
[rgeoboundaries](https://gitlab.com/dickoa/rgeoboundaries){target=blank} or 
[nominatimlite](https://github.com/dieghernan/nominatimlite){target=blank} 
for the acquisition of boundary data that can be used with 
`ohsome_boundary()`.

There are also the following methods of `ohsome_boundary()` for other classes
of input geometry objects:

1. `bbox` objects created with `st_bbox` are converted into a textual `bboxes` 
parameter to the query:

```{r bbox}
q <- ohsome_query("users/count") |>
	set_boundary(sf::st_bbox(franconia))

q$body$bboxes
```

2. `matrix` objects created with `sp::bbox()`, `raster::bbox()` or 
`terra::bbox()` are also converted into a textual `bboxes` parameter. This even 
applies for matrices created with `osmdata::getbb()` and `tmaptools::bb()`, so 
that you can comfortably acquire bounding boxes from the Nominatim API:

```{r getbb}
osmdata::getbb("Kigali") |> 
	ohsome_elements_length(time = "2018/2018-12/P1M", filter = "route=bus") |>
	ohsome_post()
```

3. You can pass any `character` object with text in the 
[format allowed by the ohsome API](https://docs.ohsome.org/ohsome-api/stable/boundaries.html){target=blank}
to `ohsome_boundary()` -- even GeoJSON FeatureCollections. It will automatically 
detect whether you have passed the definition of `bpolys`, `bboxes` or 
`bcircles`. It is possible to use `character` vectors where each element 
represents one geometry:

```{r circles}
c("Circle 1:8.6528,49.3683,1000", "Circle 2:8.7294,49.4376,1000") |>
	ohsome_elements_count(filter = "amenity=*", grouping = "boundary", time = 2021) |>
	ohsome_post()
```

`r emo::ji("warning")` While `sf` and `bbox` objects will be automatically 
transformed to WGS 84 if in a different coordinate reference system, coordinates
in `character` and `matrix` objects always need to be provided as WGS 84.

### Modifying queries

As seen above, existing `ohsome_query` objects can be modified by 
`set_endpoint()`, `set_grouping()`, `set_boundary()` or `set_time()`. The latter 
and other functions such as `set_filter()` are just wrappers around the more 
generic `set_parameters()`. This can be used to modify the parameters of a query 
in any possible way:

```{r set_parameters}
q <- ohsome_elements_count("8.5992,49.3567,8.7499,49.4371")

q |>
	set_endpoint("ratio", append = TRUE) |>
	set_parameters(
		filter = "building=*", 
		filter2 = "building=* and building:levels=*",
		time = "2010/2020/P2Y"
	) |>
	ohsome_post()

```

### Grouping

[Grouping endpoints](https://docs.ohsome.org/ohsome-api/v1/group-by.html){target=blank}
are available for aggregation resources and can be used to compute the 
aggregated results grouped by:

- boundary,
- key,
- tag, and
- type.

In many cases, a grouping by `boundary` can be combined with a grouping by `tag`. 
Some of the grouping endpoints require additional query parameters, e.g. `tag`
groupings require a `groupByKey` parameter. Not all grouping endpoints are 
available for all aggregation resources --  please consult the
[ohsome API documentation](https://docs.ohsome.org/ohsome-api/v1/group-by.html){target=blank}
for details.

You can set the `grouping` argument to any aggregation endpoint wrapper function 
(e.g. `ohsome_elements_count(grouping = c("boundary", "tag"))`) or use 
`set_grouping()` to modify existing query objects. 

### Density and ratio requests

Many 
[aggregation resources](https://docs.ohsome.org/ohsome-api/v1/endpoints.html){target=blank}
have endpoints for requesting density (i.e. count, length, perimeter or area of
features **per area**) or ratios (of OSM elements satisfying a `filter2` to elements 
satisfying a `filter`) instead of or in addition to absolute values.

You can request density or ratio values by setting the `return_value` argument
to aggregation endpoint wrapper functions (e.g. 
`ohsome_elements_count(return_value = "density")`). Mind that ratio endpoints 
require an additional `filter2` query parameter. Please consult the 
[ohsome API documentation](https://docs.ohsome.org/ohsome-api/v1/endpoints.html){target=blank}
or print `names(ohsome_endpoints)` to the console in order to check for the 
availability of specific density and ratio endpoints.

### Dealing with complex API responses

The ohsome API allows grouping aggregate values for various timestamps by 
boundary and tag at the same time. The parsed content of the response can be 
rather complex. In the following case, building feature counts for the districts 
of Franconia at two different timestamps are requested -- additionally grouped 
by the building:levels tag. To avoid lots of redundant geometries, 
comma-separated values (instead of GeoJSON) are explicitly requested as the 
response format:

```{r groupby_boundary_groupby_tag, message = FALSE}
building_levels <- franconia |>
	mutate(id  = NUTS_ID) |>
	ohsome_elements_count(grouping = c("boundary", "tag"), format = "csv") |>
	set_filter("building=* and geometry:polygon") |>
	set_time("2015/2020") |>
	set_groupByKey("building:levels") |>
	ohsome_post()

dim(building_levels)
```

The query results in a confusing data.frame with `r ncol(building_levels)` 
columns and `r nrow(building_levels)` rows. This happens because there is a 
building count column for each combination of boundary polygon and number of 
levels, while the two requested timestamps are in the rows. Fortunately, there 
is the 
[tidyr](https://tidyr.tidyverse.org){target=blank} 
package to do its magic and pivot this table into a long format 
with one value per row:

```{r tidy}
library(tidyr)

building_levels |>
	pivot_longer(-timestamp, names_to = c("id", "levels"), names_sep = "_")
```

## How to cite this package

In order to cite this package in publications, please use the citation 
information provided through `citation("ohsome")`.

-----

